---
title: "NYCDSA_R_Project"
author: "Laurel He"
output: html_document
---
# The dataset contains the list of wildfires occured in California between 2013 - 2020, the location where the wildfires occured, the County name, latitude and longitude, acres burned and other details.
```{r}
library(dplyr)
library(tidyverse)
library(sf)
library(tigris)
library(mapview)
library(leaflet)
library(tidyr)
library(lubridate)
```
```{r}
fire = read.csv('data/California_Fire_Incidents.csv')
fire
```
```{r}
myvars = c('AcresBurned', 'Counties', 'Extinguished', 'Latitude', 'Longitude', 'MajorIncident', 'Started', 'Name')
fire_clean = fire[myvars]
fire_clean = fire_clean %>% drop_na() %>% filter(!grepl('1969', Started)) %>% filter(Latitude != 0 & Longitude != 0)
fire_clean
```
```{r}
start_DT = ymd_hms(fire_clean$Started)
end_DT = ymd_hms(fire_clean$Extinguished)
duration = difftime(end_DT, start_DT, units = "hours")
new_fire = cbind(fire_clean, duration_hrs = as.numeric(duration), start_DT)
new_fire$year = year(start_DT)
new_fire = new_fire %>% drop_na() %>% filter(duration_hrs > 0 & AcresBurned > 0)
new_fire
```
```{r}
data = new_fire %>% filter(year == '2013' & Latitude == 37.85700 & Longitude == -120.0860)
data$Name
```

```{r}
ca_counties <- counties("CA", cb = T, progress_bar = F)

leaflet() %>%
  addTiles() %>%
  addCircleMarkers(
    data = new_fire %>% filter(year == '2019')
  )

```




# Wild fire frequency and severity measured by burned acreages by year
```{r}
new_fire %>% group_by(year = year(start_DT)) %>% summarise(count = n()) %>% ggplot(aes(year, count)) + geom_bar(stat = 'identity')
```
```{r}
# Major incidents count per year
new_fire %>% group_by(year = year(start_DT)) %>% filter(MajorIncident == "True") %>% summarise(count = n()) %>% ggplot(aes(year, count)) + geom_bar(stat = 'identity')
```

```{r}
new_fire %>% group_by(year = year(start_DT)) %>% summarise(totalAcres = sum(AcresBurned)) %>% ggplot(aes(year, totalAcres)) + geom_bar(stat = 'identity')
```

# Wild fire frequency and severity measured by burned acreages by time of year
```{r}
new_fire %>% group_by(month = month(start_DT)) %>% summarise(count=n()) %>% ggplot(aes(x=month, y=count)) + geom_bar(stat = 'identity')
```
```{r}
new_fire %>% group_by(month = month(start_DT)) %>% summarise(AcresBurned = sum(AcresBurned)) %>% ggplot(aes(x=month, y=AcresBurned)) + geom_bar(stat = 'identity')
```
# Wild fire frequency and severity measured by burned acreages by time of day
```{r}
new_fire %>% group_by(hour = hour(start_DT)) %>% summarise(count=n()) %>% ggplot(aes(x=hour, y=count)) + geom_bar(stat = 'identity')
```

```{r}
new_fire %>% group_by(hour = hour(start_DT)) %>% summarise(AcresBurned = sum(AcresBurned)) %>% ggplot(aes(x=hour, y=AcresBurned)) + geom_bar(stat = 'identity')
```
# total acres burned in each country from 2013 - 2020
```{r}
new_fire %>% group_by(Counties) %>% summarise(totalAcres = sum(AcresBurned)) %>% 
ggplot(aes(x=reorder(Counties, totalAcres), y=totalAcres)) + geom_bar(stat = 'identity') + coord_flip()
```
```{r}
# Boxplot of duration vs. year
new_fire %>% ggplot(aes(as.character(year(start_DT)), duration_hrs)) + geom_boxplot()
```
```{r}
fire_2018 = new_fire %>% filter(year(start_DT) == 2018) %>% filter(AcresBurned <= 100000 & duration_hrs < 7500) # focus on 2018 and remove outliers
fire_2018 %>% ggplot(aes(x=duration_hrs, y=AcresBurned)) + geom_point() 
```
```{r}
fire_2018 %>% ggplot(aes(x=duration_hrs, y=AcresBurned)) + geom_point() + scale_x_log10() + scale_y_log10()
```

```{r}
model = lm(AcresBurned ~ duration_hrs, data = fire_2018) 

bc = boxCox(model)
lambda = bc$x[which(bc$y == max(bc$y))] #Extracting the best lambda value.
fire.bc = (fire_2018$AcresBurned^lambda - 1)/lambda #Applying the Box-Cox transformation.

model.bc = lm(fire.bc ~ fire_2018$duration_hrs) #Creating a new regression based on the transformed variable.
summary(model.bc) #Assessing the output of the new model.
```
```{r}
plot(model.bc)
```


```{r}
fire_2018$MajorIncident = as.factor(fire_2018$MajorIncident) #Converting to a categorical variable.
logit.overall = glm(MajorIncident ~ AcresBurned,
                    family = "binomial",
                    data = fire_2018) # duration_hrs wasn't a significant feature
summary(logit.overall)
```
```{r}
confint(logit.overall)
```
#How does the probability of admission change across ranks for a student who has an average GRE and an average GPA?

```{r}
newdata = with(fire_2018, data.frame(AcresBurned = mean(AcresBurned),
                                       MajorIncident = factor(1:2)))
newdata #Creating a data frame with the average GRE and GPA for each level of the rank variable.
```
```{r}
cbind(newdata, "P_Major" = predict(logit.overall, newdata, type = "response"))
```
```{r}
major.predicted = round(logit.overall$fitted.values)
table(truth = fire_2018$MajorIncident, prediction = major.predicted)
```

```{r}
pchisq(logit.overall$deviance, logit.overall$df.residual, lower.tail = FALSE)
```
```{r}
1 - logit.overall$deviance/logit.overall$null.deviance
```
```{r}
table(fire_2018$MajorIncident) #Our data contains 273 unadmitted students and 127
                         #admitted students.
```
```{r}
table(major.predicted) #The model we created predicts that 351 students will
                          #not be admitted, and only 49 will be admitted.
```

```{r}
myvars = c('AcresBurned',  'MajorIncident', 'duration_hrs')
new_fire2 = new_fire[myvars]
plot(new_fire2) #Basic graphical EDA.
```

```{r}


ca_counties <- counties("CA", cb = T, progress_bar = F)
```
```{r}

```

# Shiny App to visualize fire location changes throughout the years, bubble plot: size proportional to acres burned
